---
title: "Capstone 01 Process"
author: "Tim Andaya"
date: "2025-04-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## It was necessary to create an ODBC connector for an SQL Server
## due to my preference of working with the datasets stored in 
## .csv files into a more robust application.
# ----------------------------------------------------------------------
```{r}
install.packages("odbc")
install.packages("tidyverse")

```

# ----------------------------------------------------------------------
```{r}
library(odbc)
library(tidyverse)
```
# ----------------------------------------------------------------------
```{r}
con <- dbConnect(odbc(),
                 Driver = "SQL Server",
                 Server = "TABITHA",
                 Database = "GDAC",
                 UID = "rgdac",
                 PWD = rstudioapi::askForPassword("Database password"),
                 Port = 1433)
```

# ----------------------------------------------------------------------
```{r}
# # ----------------------------------------------------------------------
# # Helper functions to run the query and report status
# # ----------------------------------------------------------------------
#' Run a non-query SQL command
#' @param con A DBI connection object
#' @param query SQL query as a string
#' @param label Optional label for logging
#' @return NULL, prints message
#' @export
run_query <- function(con, query, label = NULL) {
  tryCatch({
    result <- DBI::dbExecute(con, query)
    prefix <- if (!is.null(label)) paste0("[", label, "] ") else ""
    if (result == 0) {
      message(prefix, "Commands completed successfully.")
    } else {
      message(prefix, result, " rows affected.")
    }
  }, error = function(e) {
    message("Error: ", e$message)
  })
}

# # Execute the query (this line triggers the action)
# run_query(con, query, label = "Create divvy_trips_xxxx_qx")

#' Fetch data from a SQL query
#'
#' @param con A DBI connection object
#' @param query SQL query as a string
#' @param label Optional label for logging
#'
#' @return A data frame containing the result of the query, or NULL on error
#' @export
fetch_data <- function(con, query, label = NULL) {
  tryCatch({
    df <- DBI::dbGetQuery(con, query)
    prefix <- if (!is.null(label)) paste0("[", label, "] ") else ""
    message(prefix, "Query executed. ", nrow(df), " rows returned.")
    return(df)
  }, error = function(e) {
    message("Error: ", e$message)
    return(NULL)
  })
}

```

# ----------------------------------------------------------------------
```{r Enable python}
library(reticulate)
#install.packages("pandas") 
# remotes::install_github("pandas-dev/pandas") 
# use_virtualenv("r-reticulate")
# Method 1: Using py_run_string
# py_run_string("import pandas as pd\n# Your pandas code here")

```

# ----------------------------------------------------------------------
#Data:  Initial preparation and import into analysis and/or staging tables
# ----------------------------------------------------------------------
##Create analysis table gdac.dbo.divvy-tripdata
# ----------------------------------------------------------------------
```{r UNCOMMENT QUERY TO USE, echo=TRUE, message=TRUE, warning=TRUE}
con
query="--USE [GDAC]
--GO

--/****** Object:  Table [dbo].[divvy-tripdata]    Script Date: 4/15/2025 10:13:51 AM ******/
--IF  EXISTS (SELECT * FROM sys.objects WHERE object_id = OBJECT_ID(N'[dbo].[divvy-tripdata]') AND type in (N'U'))
--DROP TABLE [dbo].[divvy-tripdata]
--GO

--/****** Object:  Table [dbo].[divvy-tripdata]    Script Date: 4/15/2025 10:13:51 AM ******/
--SET ANSI_NULLS ON
--GO

--SET QUOTED_IDENTIFIER ON
--GO

--CREATE TABLE [dbo].[divvy-tripdata](		    --	DTS Import specification
--	[ride_id] [varchar](20) NULL,				      --	DT_STR	20				      RAW DATA
--	[rideable_type] [varchar](30) NULL,		  	--	DT_STR	30				      RAW DATA
--	[started_at] [datetime] NULL,				      --	DT_DATE					        RAW DATA
--	[ended_at] [datetime] NULL,					      --	DT_DATE					        RAW DATA
--	[start_station_name] [varchar](50) NULL,	--	DT_STR	50				      RAW DATA
--	[start_station_id] [int] NULL,				    --	Four Byte Unsigned INT	RAW DATA
--	[end_station_name] [varchar](50) NULL,		--	DT_STR	50				      RAW DATA
--	[end_station_id] [int] NULL,				      --	Four Byte Unsigned INT	RAW DATA
--	[start_lat] [decimal](8, 6) NULL,			    --	Float					          RAW DATA
--	[start_lng] [decimal](9, 6) NULL,			    --	Float					          RAW DATA
--	[end_lat] [decimal](8, 6) NULL,				    --	Float					          RAW DATA
--	[end_lng] [decimal](9, 6) NULL,				    --	Float					          RAW DATA
--	[member_casual] [varchar](10) NULL,			  --	Float					          RAW DATA
--	[ride_length] [time](7) NULL,				      --	DT_DATE					        CALCULATED 
--	[day_of_week] [int] NULL					        --	Four Byte Unsigned INT	CALCULATED
--) ON [PRIMARY]
--GO"

# Execute the query (this line triggers the action)
run_query(con, query, label = "Create table divvy-tripdata in GDAC database")
```
# ----------------------------------------------------------------------
## Create staging table divvy_trips_xxxx_qx
# ----------------------------------------------------------------------
```{r UNCOMMENT QUERY TO USE, echo=TRUE, message=TRUE, warning=TRUE}
# Database connection
con

# SQL query (currently commented out for modeling)
query = "USE [GDAC]
--GO

--/****** Object:  Table [dbo].[divvy_trips_xxxx_qx]    Script Date: 4/21/2025 11:23:08 AM ******/
--IF  EXISTS (SELECT * FROM sys.objects WHERE object_id = OBJECT_ID(N'[dbo].[divvy_trips_xxxx_qx]') AND type in (N'U'))
--DROP TABLE [dbo].[divvy_trips_xxxx_qx]
--GO

--/****** Object:  Table [dbo].[divvy_trips_xxxx_qx]    Script Date: 4/21/2025 11:23:08 AM ******/
--SET ANSI_NULLS ON
--GO

--SET QUOTED_IDENTIFIER ON
--GO

--CREATE TABLE [dbo].[divvy_trips_xxxx_qx](		    --	DTS Import specification	Comment
--	[trip_id] [varchar](20) NULL,				    --	DT_
--	[start_time] [datetime] NULL,				    --	DT_DATE					    RAW DATA
--	[end_time] [datetime] NULL,						--	DT_DATE					    RAW DATA
--	[bikeid] [varchar](50) NULL,				    --	DT_STR	50					RAW DATA
--	[tripduration] [varchar](100) NULL,				--	DT_STR	100					RAW DATA
--	[from_station_id] [varchar](100) NULL,			--	DT_STR	100					RAW DATA
--	[from_station_name] [varchar](100) NULL,		--	DT_STR	100					RAW DATA
--	[to_station_id] [varchar](100) NULL,			--	DT_STR	100					RAW DATA
--	[to_station_name] [varchar](100) NULL,			--	DT_STR	100					RAW DATA
--	[usertype] [varchar](100) NULL,					--	DT_STR	100					RAW DATA
--	[gender] [varchar](50) NULL,				    --	DT_STR	50					RAW DATA
--	[birthyear] [varchar](50) NULL,				    --	DT_STR	50					RAW DATA
--	[ride_length] [varchar](50) NULL,				--	DT_STR	50					CALCULATED
--	[day_of_week] [varchar](50) NULL				--	DT_STR	50					CALCULATED
--) ON [PRIMARY]
--GO"

# Execute the query (this line triggers the action)
run_query(con, query, label = "Create divvy_trips_xxxx_qx")

```

 NOTES:
      Calculations will need to get adjusted for the non 
      xxxx-divvy-tripdata files as/if necessary and will be handled 
      after the files have been imported into either the main analysis
      database, gdac.dbo.divvy-tripdata, or the staging table created 
      for the divvy_trips_YYYY_QX files format  
      Problematic files have commas in the data where numeric values exceed
      999.99, a reason why I use pipes as a delimiter when I export data; 
      I created a python script to address this in all data sets.  
      Additionally, this script will add the two new column headings and 
      corresponding commas at the data level and finally, confirm that 
      each line has a hard return.
  
##Python was used to interate trough the files.
*Add two column names to the header; ride_length & day_of_week.
*Add two commas on each data line.
*Add a hard line return to the end of each line.
*Each modified file's content was added to a new file.

##Python first stage data cleanup file set 1

```{python Modify xxx-divvy-tripdata CSV Files}
import csv
import re
import os
from pathlib import Path

def clean_numeric_field(value):
    # Remove commas from numeric strings (both quoted and unquoted)
    if re.fullmatch(r'\d{1,3}(,\d{3})*(\.\d+)?', value):
        return value.replace(',', '')
    return value

def process_csv(input_file, output_file):
    if not os.path.exists(input_file):
        print(f"Input file not found: {input_file}")
        return

    with open(input_file, 'r', encoding='utf-8', newline='') as infile, \
         open(output_file, 'w', encoding='utf-8', newline='\n') as outfile:

        reader = csv.reader(infile, delimiter=',', quotechar='"')
        writer = csv.writer(outfile, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)

        headers = next(reader)
        headers += ['ride_length', 'day_of_week']
        writer.writerow(headers)

        line_count = 0
        for row in reader:
            cleaned_row = []
            for field in row:
                cleaned = clean_numeric_field(field)
                # Uncomment this line to debug field changes:
                # if cleaned != field:
                #     print(f"Changed: {field} -> {cleaned}")
                cleaned_row.append(cleaned)
            cleaned_row += ['', '']
            writer.writerow(cleaned_row)
            line_count += 1

        print(f"Processed {line_count} rows. Output saved to: {output_file}")

# === Batch Processing ===

input_dir = Path(r'\\tabitha\gdac$\CS1')
output_dir = input_dir  # Or change to another Path if needed

# List of input files to process
file_names = [
"202004-divvy-tripdata.csv",
"202005-divvy-tripdata.csv",
"202006-divvy-tripdata.csv",
"202007-divvy-tripdata.csv",
"202008-divvy-tripdata.csv",
"202009-divvy-tripdata.csv",
"202010-divvy-tripdata.csv",
"202011-divvy-tripdata.csv",
"202012-divvy-tripdata.csv",
"202101-divvy-tripdata.csv",
"202102-divvy-tripdata.csv",
"202103-divvy-tripdata.csv",
"202104-divvy-tripdata.csv",
"202105-divvy-tripdata.csv",
"202106-divvy-tripdata.csv",
"202107-divvy-tripdata.csv",
"202108-divvy-tripdata.csv",
"202109-divvy-tripdata.csv",
"202110-divvy-tripdata.csv",
"202111-divvy-tripdata.csv",
"202112-divvy-tripdata.csv",
"202201-divvy-tripdata.csv",
"202202-divvy-tripdata.csv",
"202203-divvy-tripdata.csv",
"202204-divvy-tripdata.csv",
"202205-divvy-tripdata.csv",
"202206-divvy-tripdata.csv",
"202207-divvy-tripdata.csv",
"202208-divvy-tripdata.csv",
"202209-divvy-publictripdata.csv",
"202210-divvy-tripdata.csv",
"202211-divvy-tripdata.csv",
"202212-divvy-tripdata.csv",
"202301-divvy-tripdata.csv",
"202302-divvy-tripdata.csv",
"202303-divvy-tripdata.csv",
"202304-divvy-tripdata.csv",
"202305-divvy-tripdata.csv",
"202306-divvy-tripdata.csv",
"202307-divvy-tripdata.csv",
"202308-divvy-tripdata.csv",
"202309-divvy-tripdata.csv",
"202310-divvy-tripdata.csv",
"202311-divvy-tripdata.csv",
"202312-divvy-tripdata.csv",
"202401-divvy-tripdata.csv",
"202402-divvy-tripdata.csv",
"202403-divvy-tripdata.csv",
"202404-divvy-tripdata.csv",
"202405-divvy-tripdata.csv",
"202406-divvy-tripdata.csv",
"202407-divvy-tripdata.csv",
"202408-divvy-tripdata.csv",
"202409-divvy-tripdata.csv",
"202410-divvy-tripdata.csv",
"202411-divvy-tripdata.csv",
"202412-divvy-tripdata.csv",
"202501-divvy-tripdata.csv",
"202502-divvy-tripdata.csv",
"202503-divvy-tripdata.csv",
"Divvy_Trips_2020_Q1.csv"
]

# Process each file
for file_name in file_names:
    input_path = input_dir / file_name
    output_name = file_name.replace(".csv", "_PyTgt.csv")
    output_path = output_dir / output_name
    process_csv(str(input_path), str(output_path))

```

# ----------------------------------------------------------------------
## Python first stage data cleanup file set 2
# ----------------------------------------------------------------------
```{python Modify divvy-trips_xxxx_qx CSV Files}
import csv
import re
import os
from pathlib import Path

def clean_numeric_field(value):
    # Remove commas from numeric strings (e.g., 1,783.0 → 1783.0)
    if re.fullmatch(r'\d{1,3}(,\d{3})*(\.\d+)?', value):
        return value.replace(',', '')
    return value

def process_csv(input_file, output_file):
    if not os.path.exists(input_file):
        print(f"Input file not found: {input_file}")
        return

    with open(input_file, 'r', encoding='utf-8', newline='') as infile, \
         open(output_file, 'w', encoding='utf-8', newline='\n') as outfile:

        reader = csv.reader(infile, delimiter=',', quotechar='"')
        writer = csv.writer(outfile, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)

        headers = next(reader)
        headers += ['ride_length', 'day_of_week']
        writer.writerow(headers)

        line_count = 0
        for row in reader:
            cleaned_row = []
            for field in row:
                cleaned = clean_numeric_field(field)
                # Uncomment this line to debug field changes:
                # if cleaned != field:
                #     print(f"Changed: {field} -> {cleaned}")
                cleaned_row.append(cleaned)
            cleaned_row += ['', '']
            writer.writerow(cleaned_row)
            line_count += 1

        print(f"Processed {line_count} rows. Output saved to: {output_file}")

# === Batch Processing ===

input_dir = Path(r'\\tabitha\gdac$\CS1')
output_dir = input_dir  # Or change to another Path if needed

file_names = [
    "Divvy_Trips_2019_Q1.csv",
    "Divvy_Trips_2019_Q2.csv",
    "Divvy_Trips_2019_Q3.csv",
    "Divvy_Trips_2019_Q4.csv"
]

for file_name in file_names:
    input_path = input_dir / file_name
    output_name = file_name.replace(".csv", "_PyTgt.csv")
    output_path = output_dir / output_name
    process_csv(str(input_path), str(output_path))
```

# ----------------------------------------------------------------------
##Load files into tables
# ----------------------------------------------------------------------
I created a cursor script to bulk load the data, I found problems with 
  the data that were collected near the end of 2024 to March of 2025; 
  it would seem that some form of collection paradigm shift occurred 
  in June of 2024

###divvy-tripdata records
    CSV files with YYYYMM-divvy-tripdata naming convention that have valid
    dates import into the gdac.dbo.divvy-tripdata table without issue.

##divvy_Trips_YYYY_Qx 
    Files with the naming convention Divvy_Trips_ must be loaded into a 
    staging table, transformed prior to insertion into 
    gdac.dbo.divvy-trips_xxxx_qx after they have been modified using 
    python to add two new columns, ride_length and day_of_week, stripping
    out commas from numerical values encapsulated by double quotes, and 
    adding two commas at the end of each data line to create the new
    field data placeholders.
    Again, while I could manually import the data, this was inefficient; I 
    needed a quicker and more streamline method.  On top of that, Excel
    could not open the larger quarterly files for manipulation without 
    data loss.

#### divvy_Trips_YYYY_Qx BCP Format File
    A format file for the quarterly tables that matched the
    staging table was required to load these files

    <?xml version="1.0"?>
    <BCPFORMAT xmlns="http://schemas.microsoft.com/sqlserver/2004/bulkload/format" xmlns:xsi="http://www.w3.org/2001/XMLSchema-instance">
     <RECORD>
      <FIELD ID="1" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="20" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="2" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="24"/>
      <FIELD ID="3" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="24"/>
      <FIELD ID="4" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="50" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="5" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="6" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="7" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="8" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="9" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="10" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="100" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="11" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="50" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="12" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="50" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="13" xsi:type="CharTerm" TERMINATOR="," MAX_LENGTH="50" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
      <FIELD ID="14" xsi:type="CharTerm" TERMINATOR="\r\n" MAX_LENGTH="50" 
        COLLATION="SQL_Latin1_General_CP1_CS_AS"/>
     </RECORD>
     <ROW>
      <COLUMN SOURCE="1" NAME="trip_id" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="2" NAME="start_time" xsi:type="SQLDATETIME"/>
      <COLUMN SOURCE="3" NAME="end_time" xsi:type="SQLDATETIME"/>
      <COLUMN SOURCE="4" NAME="bikeid" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="5" NAME="tripduration" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="6" NAME="from_station_id" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="7" NAME="from_station_name" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="8" NAME="to_station_id" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="9" NAME="to_station_name" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="10" NAME="usertype" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="11" NAME="gender" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="12" NAME="birthyear" xsi:type="SQLVARYCHAR"/>
      <COLUMN SOURCE="13" NAME="ride_length" xsi:type="SQLVARYCHAR" NULLABLE="YES"/>
      <COLUMN SOURCE="14" NAME="day_of_week" xsi:type="SQLVARYCHAR" NULLABLE="YES"/>
     </ROW>
    </BCPFORMAT>

# ----------------------------------------------------------------------
## Load gdac.dbo.divvy-tripdata table
# ----------------------------------------------------------------------
```{r SQL CSV load stored procedure}
con
query = "--exec up_load_divvy @task = 0"

run_query(con, query, "BULK LOAD CSV DATA")

```
##Update calculated fields

```{r SQL calculate ride length and day of the week values}
con
query = "
UPDATE dtxq
set ride_length = CONVERT(char(10),end_time-start_time, 108)
	-- CAST(CAST(ROUND(dbo.invoice_line.qty_shipped, 0) as int) as varchar(40))
	,day_of_week =CAST(CAST(ROUND(DATEPART(WEEKDAY,start_time), 0) AS INT) AS VARCHAR(1))
FROM [dbo].[divvy_trips_xxxx_qx] dtxq"

run_query(con,query,"Update divvy_trips_xxxx_qx ride length and day of week values")
```
  
```{r SQL calculate ride length and day of the week values}
con
query = "
UPDATE dt
set ride_length = CONVERT(char(10),ended_at-started_at, 108)
	-- CAST(CAST(ROUND(dbo.invoice_line.qty_shipped, 0) as int) as varchar(40))
	,day_of_week =CAST(CAST(ROUND(DATEPART(WEEKDAY,started_at), 0) AS INT) AS VARCHAR(1))
FROM [dbo].[divvy-tripdata] dt"

run_query(con,query,"Update divvy_trips_xxxx_qx ride length and day of week values")
```
##Data differences between standard tripdata and trip_xxxx_qx
  The Quarterly data for 2019 used user type identitifier "Customer"
  and "Subscribber", the rest of the data tables, including the
  trips_2020_Q1 data used "casual" and "member".
  The 2019 data in the trip_xxxx_qx table was updated to change the
  records to replace "Customer" with "casual", and "Subscriber" with
  "member".
```{r update divvy_trips_xxxx_qx usertype data Customer}
con
query = "UPDATE dtxq
SET usertype = 'casual'
FROM [dbo].[divvy_trips_xxxx_qx] dtxq where usertype = 'Customer'"

run_query(con,query,"Update divvy_trips_xxxx_qx usertype Customer to casual")
```
  
```{r update divvy_trips_xxxx_qx usertype data Subscriber}
con
query = "UPDATE dtxq
SET usertype = 'member'
FROM [dbo].[divvy_trips_xxxx_qx] dtxq where usertype = 'Subscriber'"

run_query(con,query,"Update divvy_trips_xxxx_qx usertype Subscriber to member")
```
## Insert divvy_xxxx_qx into divvy-tripdata
```{r}
con
query = "--INSERT INTO [dbo].[divvy-tripdata](
--	[ride_id]
--	,[started_at]
--	,[ended_at]
--	,[rideable_type]
--	,[start_station_name]
--	,[start_station_id]
--	,[end_station_name]
--	,[end_station_id]
--	,[start_lat]
--	,[start_lng]
--	,[end_lat]
--	,[end_lng]
--	,[member_casual]
--	,[ride_length]
--	,[day_of_week]
--	  )
--SELECT 
--	[trip_id] AS ride_id
--	,[start_time]
--	,[end_time]
--	,'' AS rideable_type
--	,[from_station_name] AS stat_station_name
--	,[from_station_id] AS start_station_id
--	,[to_station_name] AS end_station_name
--	,[to_station_id] AS end_station_id
--	,0 AS start_lat
--	,0 AS start_lng
--	,0 AS end_lat
--	,0 AS end_lng
--	,[usertype] AS member_casual
--	,CONVERT(char(10),end_time-start_time, 108) AS ride_length
--	,CAST(CAST(ROUND(DATEPART(WEEKDAY,start_time), 0) AS INT) AS VARCHAR(1)) AS day_of_week
--  FROM [GDAC].[dbo].[divvy_trips_xxxx_qx]
-- -- (3818004 rows affected)
-- -- Completion time: 2025-04-24T10:49:10.9604695-07:00
"
# 
run_query(con,query,"Insert divvy_xxxx_qx into divvy-tripdata")
```
##Create analysis table of 2019-2020 data summarizing ride data 
  By 
  1. period_yr = YYYY [INT]
  2. period_mo = MM   [INT]
  3. cutomer_typ = member_casual
  4. dow = day_of_week
  5. time_of_day = caluculated morning, afternoon, or evening
  6. ride_cou = count of ride_id[s] in the range
  7. avg_ride_length = the average ride length in the range

###Drop div_py_pm_dow_tod_2019_2020 table if it exists
```{r Drop div_py_pm_dow_tod_2019_2020 if it exists}
con
query="-- IF  EXISTS (SELECT * FROM sys.objects WHERE object_id = OBJECT_ID(N'[dbo].[div_py_pm_dow_tod_2019_2020]') AND type in (N'U'))
-- DROP TABLE [dbo].[div_py_pm_dow_tod_2019_2020]
"
# 
run_query(con,query,"Drop div_py_pm_dow_tod_2019_2020")
```

###Create div_py_pm_dow_tod_2019_2020 table
```{r}
con
query="-- CREATE TABLE [dbo].[div_py_pm_dow_tod_2019_2020](
	-- [period_yr] [int] NULL,
	-- [period_mo] [int] NULL,
	-- [customer_type] [varchar](10) NULL,
	-- [dow] [varchar](3) NULL,
	-- [time_of_day] [varchar](9) NOT NULL,
	-- [ride_count] [int] NULL,
	-- [avg_ride_length] [int] NULL
-- ) ON [PRIMARY]"
# 
run_query(con,query,"Create table div_py_pm_dow_tod_2019_2020")
```
  
###Load div_py_pm_dow_tod_2019_2020 table
```{r}
con
query="-- INSERT INTO [dbo].[div_py_pm_dow_tod_2019_2020]
-- SELECT
--     DATEPART(YEAR, td.started_at) AS period_yr,
--     DATEPART(MONTH, td.started_at) AS period_mo,
--     td.member_casual AS customer_type,
--     CASE td.day_of_week
--         WHEN 1 THEN 'Sun'
--         WHEN 2 THEN 'Mon'
--         WHEN 3 THEN 'Tue'
--         WHEN 4 THEN 'Wed'
--         WHEN 5 THEN 'Thu'
--         WHEN 6 THEN 'Fri'
--         WHEN 7 THEN 'Sat'
--     END AS dow,
--     CASE 
--         WHEN CONVERT(time, td.started_at) >= '05:00:00' AND CONVERT(time, td.started_at) < '12:00:00' THEN 'morning'
--         WHEN CONVERT(time, td.started_at) >= '12:00:00' AND CONVERT(time, td.started_at) < '17:00:00' THEN 'afternoon'
--         ELSE 'evening'
--     END AS time_of_day,
--     COUNT(td.ride_id) AS ride_count,
--     AVG(DATEDIFF(MINUTE, td.started_at, td.ended_at)) AS avg_ride_length
-- --INTO div_py_pm_dow_tod_2019_2020
-- FROM dbo.[divvy-tripdata] td WITH (INDEX(div_stat_mc_dow))
-- WHERE DATEPART(YEAR, td.started_at) IN (2019, 2020)
-- GROUP BY 
--     DATEPART(YEAR, td.started_at),
--     DATEPART(MONTH, td.started_at),
--     td.member_casual,
--     td.day_of_week,
--     CASE 
--         WHEN CONVERT(time, td.started_at) >= '05:00:00' AND CONVERT(time, td.started_at) < '12:00:00' THEN 'morning'
--         WHEN CONVERT(time, td.started_at) >= '12:00:00' AND CONVERT(time, td.started_at) < '17:00:00' THEN 'afternoon'
--         ELSE 'evening'
--     END
-- ORDER BY period_yr, period_mo, td.day_of_week, time_of_day;"
# 
run_query(con,query,"Load div_py_pm_dow_tod_2019_2020")
```

### Load data frame div_pid_ct_dow_tod_rc_arl
```{r load data frame div_pid_ct_dow_tod_rc_arl}
con
query="SELECT
  CAST([period_yr] AS varchar(4)) AS period_year,
  CAST([period_yr] AS varchar(4)) + RIGHT('0' + CAST([period_mo] AS varchar(2)), 2) AS period_id,
  [customer_type],
  [dow],
  [time_of_day],
  [ride_count],
  [avg_ride_length]
FROM [GDAC].[dbo].[div_py_pm_dow_tod_2019_2020]
ORDER BY 
  period_yr, 
  period_mo,
  CASE dow 
    WHEN 'Sun' THEN 1 
    WHEN 'Mon' THEN 2 
    WHEN 'Tue' THEN 3  
    WHEN 'Wed' THEN 4 
    WHEN 'Thu' THEN 5 
    WHEN 'Fri' THEN 6 
    ELSE 7 
  END,
  customer_type,
  CASE time_of_day 
    WHEN 'morning' THEN 1 
    WHEN 'afternoon' THEN 2 
    ELSE 3 
  END"
#
div_pid_ct_dow_tod_rc_arl <- fetch_data(con,query)
```


## Visualizations

Here, we will go through a series of visualizations

#### *Ridership by period*
```{r}
ggplot(data=div_pid_ct_dow_tod_rc_arl,aes(x=period_id,y=ride_count))+
  geom_point(aes(shape=customer_type, color= customer_type))+
  facet_wrap(~customer_type)+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position="none") +
  labs(title = "Casual use increased in 2020 but member use decreased.  Peak use summer",
       x = "Periods",
       y = "Rider Counts")


```
# Create a bar plot with vertical x-axis labels
```{r}

ggplot(div_pid_ct_dow_tod_rc_arl, aes(x = period_id, y = ride_count, fill=ride_count)) +
  geom_bar(stat = "Identity")+scale_fill_gradient(low="blue",high="red") +
  facet_wrap(~customer_type)+
  # theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position="none") +
  labs(title = "Cas. use increased in 2020 but mbr use decreased.  Usage peak is summer",
       x = "Periods",
       y = "Rider Counts")
```


#### *Weekly ridership across 2019-2020*
```{r}
## this vector is for weekday sorting
level_weekdays <- c('Sun', 'Mon','Tue', 'Wed', 'Thu', 'Fri', 'Sat')
## Plot
ggplot(data=div_pid_ct_dow_tod_rc_arl,aes(x=factor(dow, level = level_weekdays),y=ride_count))+
  geom_point(aes(shape=customer_type, color= customer_type))+
  facet_grid(customer_type ~ period_year)+
   theme_minimal()+ 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),legend.position="none") +
  labs(title = "Cas. use higher Friday through Sunday across 2019-2020",
       x = "Periods",
       y = "Rider Counts")
```
