---
title: "Capstone 1Divvy Charts"
author: "Tim Andaya"
date: "2025-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## It was necessary to create an ODBC connector for an SQL Server
## due to my preference of working with the datasets stored in 
## .csv files into a more robust application.
# ----------------------------------------------------------------------
Packages loaded
*odbc
*tidyverse
```{r}
# install.packages("odbc")
# install.packages("tidyverse")

```

# ----------------------------------------------------------------------
Libraries  and functions used
*odbc
*tidyvers

```{r}
library(odbc)
library(tidyverse)
```
# ----------------------------------------------------------------------

```{r}
con <- dbConnect(odbc(),
                 Driver = "SQL Server",
                 Server = "TABITHA",
                 Database = "GDAC",
                 UID = "rgdac",
                 PWD = rstudioapi::askForPassword("Database password"),
                 Port = 1433)
```

*run_query
*fectch_data

```{r}
# # ----------------------------------------------------------------------
# # Helper functions to run the query and report status
# # ----------------------------------------------------------------------
#' Run a non-query SQL command
#' @param con A DBI connection object
#' @param query SQL query as a string
#' @param label Optional label for logging
#' @return NULL, prints message
#' @export
run_query <- function(con, query, label = NULL) {
  tryCatch({
    result <- DBI::dbExecute(con, query)
    prefix <- if (!is.null(label)) paste0("[", label, "] ") else ""
    if (result == 0) {
      message(prefix, "Commands completed successfully.")
    } else {
      message(prefix, result, " rows affected.")
    }
  }, error = function(e) {
    message("Error: ", e$message)
  })
}

# # Execute the query (this line triggers the action)
# run_query(con, query, label = "Create divvy_trips_xxxx_qx")

#' Fetch data from a SQL query
#'
#' @param con A DBI connection object
#' @param query SQL query as a string
#' @param label Optional label for logging
#'
#' @return A data frame containing the result of the query, or NULL on error
#' @export
fetch_data <- function(con, query, label = NULL) {
  tryCatch({
    df <- DBI::dbGetQuery(con, query)
    prefix <- if (!is.null(label)) paste0("[", label, "] ") else ""
    message(prefix, "Query executed. ", nrow(df), " rows returned.")
    return(df)
  }, error = function(e) {
    message("Error: ", e$message)
    return(NULL)
  })
}

```

# ----------------------------------------------------------------------
### Load data frame div_pid_ct_dow_tod_rc_arl
This is taken from a subset of the tripdata data set provided for this
  project.

```{r load data frame div_pid_ct_dow_tod_rc_arl}
con
query="SELECT
  CAST([period_yr] AS varchar(4)) AS period_year,
  CAST([period_yr] AS varchar(4)) + RIGHT('0' + CAST([period_mo] AS varchar(2)), 2) AS period_id,
  [customer_type],
  [dow],
  [time_of_day],
  [ride_count],
  [avg_ride_length]
FROM [GDAC].[dbo].[div_py_pm_dow_tod_2019_2020]
ORDER BY 
  period_yr, 
  period_mo,
  CASE dow 
    WHEN 'Sun' THEN 1 
    WHEN 'Mon' THEN 2 
    WHEN 'Tue' THEN 3  
    WHEN 'Wed' THEN 4 
    WHEN 'Thu' THEN 5 
    WHEN 'Fri' THEN 6 
    ELSE 7 
  END,
  customer_type,
  CASE time_of_day 
    WHEN 'morning' THEN 1 
    WHEN 'afternoon' THEN 2 
    ELSE 3 
  END"
#
div_pid_ct_dow_tod_rc_arl <- fetch_data(con,query)
div_pid_ct_dow_tod_rc_arl
```

# ----------------------------------------------------------------------
# Create Visualizations
# ----------------------------------------------------------------------
This chart shows that casual ridership increased in 2020 over 2019
  and that ridership decreased for member users during 2020.
This chart also shows that ridership is higher during the warmer months 
  of the year from May through September/October.
##* *2019-2020 Monthly Ridership Chart One*

```{r}
ggplot(data=div_pid_ct_dow_tod_rc_arl,aes(x=period_id,y=ride_count))+
  geom_point(aes(shape=customer_type, color= customer_type))+
  facet_wrap(~customer_type)+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position="none") +
  labs(title = "Casual use increased in 2020 but member use decreased.  Peak use summer",
       x = "Periods",
       y = "Rider Counts")

```

# ----------------------------------------------------------------------
##*2019-2020 Monthly Ridership Chart Two*
This is an alternate view of 2019-2020 ridership data.

```{r}

ggplot(div_pid_ct_dow_tod_rc_arl, aes(x = period_id, y = ride_count, fill=ride_count)) +
  geom_bar(stat = "Identity")+scale_fill_gradient(low="blue",high="red") +
  facet_wrap(~customer_type)+
  # theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position="none") +
  labs(title = "Cas. use increased in 2020 but mbr use decreased.  Usage peak is summer",
       x = "Periods",
       y = "Rider Counts")
```

# ----------------------------------------------------------------------
##*Weekly ridership across 2019-2020*
This chart shows the weekly ridership tendencies of the two types of 
  riders.  Fridays, Saturdays, and Sundays are the most popular days for
  casual riders and Monday, Tuesday, Wednesday and Thursday are the most
  popular for the members.
  Ridership for members is down in 2020

```{r}
## this vector is for weekday sorting
level_weekdays <- c('Sun', 'Mon','Tue', 'Wed', 'Thu', 'Fri', 'Sat')
## Plot
ggplot(data=div_pid_ct_dow_tod_rc_arl,aes(x=factor(dow, level = level_weekdays),y=ride_count))+
  geom_point(aes(shape=customer_type, color= customer_type))+
  facet_grid(customer_type ~ period_year)+
   theme_minimal()+ 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),legend.position="none") +
  labs(title = "Cas. use higher Friday through Sunday across 2019-2020",
       x = "Periods",
       y = "Rider Counts")
```
